---
title: "Week 1 seminar"
author: "Anastasia Chanbour"
date: "01/10/2021"
output: html_document
---

**Aim**: Visualising the gapminder data.

#  Task 1: Gapminder analysis

##  1.1: Loading the required packages
```{r}
library("gapminder")
library("tidyverse")
```

##  1.2
Generating a scatterplot with x=gdpPercap, and y=lifeExp

 - we call the data object from 2002 gm2002
 - we call the scatterplot gm_scatterplot
 - we display the plot


```{r}
gm2002 <- gapminder %>% filter(year == 2002)
gm_scatterplot <- ggplot(data=gm2002, aes(x=gdpPercap, y=lifeExp)) +
  geom_point()+
  labs(title="Scatter plot")+
  theme_bw()

gm_scatterplot

```


##  1.3: Adding lines to the scatterplot to show the linear model and loess model

```{r}
model_lm <- lm(lifeExp ~ gdpPercap, data = gm2002)
predictions_lm <- broom::augment(model_lm)
gm_scatterplot + geom_line(data = predictions_lm, aes(y=.fitted))

```

## 1.3.2 loess model to predict lifeExp

```{r}
model_loess <- loess(lifeExp ~ gdpPercap,
                     data = gm2002,
                     span = 0.1)

predictions_loess <- broom::augment(model_loess)
summary(predictions_loess)

```

Now all three models together
```{r}

gm_scatterplot + geom_line(data = predictions_lm, aes(y=.fitted), color = "blue", linetype = "dashed") + geom_line(data = predictions_loess, aes(y=.fitted), color = "green")
```
